---
title: "Estimating the number Pi by Monte Carlo Method"
author: "patmau"
date: "Nov 1, 2020"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## The Number $\pi$

The number $\pi$ is a mathematical constant and is defined as the ratio of a circle's circumference to its diameter.

It is an irrational number, i.e., $\pi$ cannot be represented as a common fraction.

See [Wikipedia](https://en.wikipedia.org/wiki/Pi) for a lot of information about this fascinating number.

### Methods for estimating $\pi$

- [Geometrical approach using polygons](https://en.wikipedia.org/wiki/Pi#Polygon_approximation_era), pioneered by Archimedes

- Infinite series

- Monte Carlo methods

## Monte Carlo method

Draw a circle inscribed in a square, then place random dots inside the square.

```{r plot, echo=FALSE, fig.height=4, fig.align='center'}
library(plotrix)

x <- runif(500, -1, 1)
y <- runif(500, -1 ,1)

plot(x, y, 
     asp = 1, 
     xlim = c(-1, 1), 
     ylim = c(-1, 1),
     xlab = "x", ylab  = "y",
     col = "darkgrey", pch = 4)
rect(-1, -1, 1, 1, border = "red", lwd = 3)
draw.circle(0, 0, 1, border = "blue", lwd = 3)
```

Assuming a circle with radius $r = 1$, the circle's area is $\pi * r^2 = \pi$, and the square's area is $(2*r)^2 = 4$.

Thus, the ratio of points within the circle to the total number of points estimates the number $\pi / 4$.

## Better estimate with more points?

The code below estimates $\pi$ for an increasing number of random dots.

```{r series, echo = TRUE}
set.seed(999)

exponents <- 2:6
pi <- sapply(exponents, function(n) {
    x <- runif(10^n, -1, 1)
    y <- runif(10^n, -1, 1)
    4 * mean(x^2 + y^2 <= 1)
})
cbind(n = 10^exponents, pi)
```

## Web application

[Here](https://patmau.shinyapps.io/EstimatingPi) is a shiny app where you can estimate $\pi$ for any number of dots between 10 and 10000 (upper limit of 10000 for performance reasons).
